### 2020-07-18 article.tex
Fixing some red part

The paper is divided in two main parts {\color{red} COMPLETE WITH MORE DETAILS WHEN PAPER IS COMPLETED}:
\cref{sec:KSPHPDDM} introduces \pk{KSPHPDDM}, provides applications within an eigensolver
framework and discuss the readiness of the PETSc library in delivering highly efficient block Krylov methods.
\cref{sec:PCHPDDM} introduces \pk{PCHPDDM}. Eventually, concluding remarks may be found
\cref{sec:conclusion}. All PETSc keywords and options are typeset in a \texttt{typewriter} font, these
are public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.

@@

\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is automatically
registered since PETSc version 3.12 when configuring PETSc with the extra
flag \pk{-{}-download-hpddm}. HPDDM linear solvers can be selected using the command
line option \pk{-ksp\_type hpddm}, or the routine \pk{KSPSetType}(ksp,
\pk{KSPHPDDM}), and the following Krylov methods can be accessed:
\begin{itemize}
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible}, with deflation at each restart;
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown}, with deflation at each iteration;
    \item block GCRODR or flexible GCRODR, with deflation at each restart.
\end{itemize}
While being mathematically equivalent to
their ``standard'' counterparts, the pseudo-block variants try to fuse together multiple similar
operations like matrix--vector
products to achieve higher arithmetic intensity, or to decrease the
number of global synchronizations needed for scalar products.
HPDDM Krylov methods may be selected {\color{red} TODO API FOR KSPHPDDMSetVariant?} using the
options \pk{-ksp\_hpddm\_type}
\pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)}, or the
routine \pk{KSPHPDDMSetType}(ksp, \pk{KSPHPDDMType}).
Preconditioning variants can be specified via
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether the
preconditioner is applied on the left, on the right, or if it cannot be
represented as a linear operator. In this latter case, the preconditioner is
always applied on the right, except for FCG. HPDDM CG
and its variants BCG and BFBCG only handle left-preconditioning.  It is also
possible to set the preconditioning side through the more common PETSc option
\pk{-ksp\_pc\_side (left$|$right)}.  Finally, convergence monitoring with the
\pk{KSPMonitor} interface is fully supported, as well as the specification of
customized convergence testing via the \pk{KSPSetConvergenceTest} callback.

According to the standard PETSc notation, ksp($A$, $P_{A'}$)
defines a \pk{KSP} used to solve a system $Ax=b$ using a preconditioner $P_{A'}$ built
using an operator $A'$ which is usually, but not necessarily, equal to $A$. While solving
the linear system via \pk{KSPSolve}(ksp, $b$, $x$) or \pk{KSPMatSolve}(ksp, $B$, $X$),
HPDDM will repeatedly call the following PETSc routines:
\begin{itemize}
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, {\small \pk{MAT\_REUSE\_MATRIX}}, {\small
        \pk{PETSC\_DEFAULT}}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}($P_{A'}$, $x$, $y$) for $y = P_{A'}x$;
    \item \pk{PCMatApply}($P_{A'}$, $X$, $Y$) for $Y = P_{A'}X$;
\end{itemize}
As of version 3.14.0, PETSc provides unified routines for applying $A$ or $P_{A'}$
to a tall-and-skinny dense matrix $X$. However, when no specialized implementation
is available, PETSc will loop over the columns of $X$ and successively apply $A$
or $P_{A'}$ to get all columns of $Y$. For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}:
\begin{itemize}
    \item \pk{MatSeqAIJ}, standard sequential sparse matrix, based on compressed sparse row format;
    \item \pk{MatSeqBAIJ}, block sparse matrix, based on block compressed sparse row format;
    \item \pk{MatSeqSBAIJ}, symmetric block sparse matrix stored in upper triangular form.
    \item \pk{MatMPIAIJ}, standard parallel sparse matrix;
    \item \pk{MatShell}, user-defined matrix;
    \item \pk{MatNest}, block-defined matrix with nested submatrices;
    \item \pk{MatSeqAIJCUSPARSE}, standard sequential sparse matrix, offloaded to a GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}
The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    \item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: embedded Krylov method inside $P_{A'}$;
    \item \pk{PCMAT}: multiplication by $A'$, i.e., $P_{A'} = {A'}$;
    \item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    % \item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}~\cite{10.1145/3232850};
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization,
        e.g., $P_{A'} = {U}^{-1}{L}^{-1}$;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like preconditioners 
\cref{it:hpddm,it:asm,it:bjacobi} is that they most often rely on exact or inexact
factorizations~\cref{it:lu,it:ilu} as subdomain solvers.
In these cases, it is possible to access the so-called local or distributed factored matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to take advantage
of blocked forward eliminations and backward substitutions from the various factorization packages
interfaced with PETSc. In the case of a sparse coefficient matrix $A'$,
this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL
PARDISO or CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense coefficient
matrix $A'$: ScaLAPACK and Elemental~\cite{poulson2013elemental}. 
Future work will consider extending the various implementations of PETSc
preconditioning infrastructure to increase the arithmetic intensity of the
block preconditioner application phase.

The rest of the operations are then performed directly inside HPDDM.
Specifically, within Krylov methods using the Arnoldi process, or when recycling is requested,
one has to compute $QR$ factorizations of tall-and-skinny dense matrices to orthonormalize
candidate basis vectors. Such operations are performed using the CholQR algorithm~\cite{stathopoulos2002block}, or
via the (modified) Gram--Schmidt method.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.
These orthonormalization variants can be selected at runtime with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}. {\color{red} API?}

@@

With these added functions, it is now possible to do code refactorization for
applications that deal with multiple right-hand sides. This is the case of some
eigensolvers in SLEPc~\cite{hernandez2005ssf}, the Scalable Library for
Eigenvalue Problem computations. SLEPc is an add-on library that extends PETSc
with several classes for linear and nonlinear eigenvalue problems, as well as
tools to, e.g., facilitate the use of shift-and-invert spectral transformations.
The implementations of LOBPCG~\cite{knyazev2001toward} and CISS~\cite{sakurai2003projection}
in SLEPc were previously not using blocking when applying the preconditioner or solving linear systems.
In version 3.14.0 we have adapted these solvers to employ \pk{KSPMatSolve} and \pk{PCMatApply}.
In the case of LOBPCG, which is a purely blocked method, all steps were already
implemented as block operations, except the applications of the preconditioner,
which ruined the benefits of blocking. With the new implementation, the gains
are striking, as reported in~\cref{sec:lobpcg}.

@@

A similar speedup can be observed in terms of runtime, with the time spent for the solution of the eigenvalue problem being reduced from \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1509}{\sec} to \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1052}{\sec}. {\color{red} FIXME LATER: These timings, and all which follow except in~\cref{sec:primitives}, have been obtained on {Ir\`ene}, a system composed of \pgfmathprintnumber[assume math mode=true]{1656} nodes with two 24-core Intel Xeon Platinum 8168 clocked at \SI{2.7}{\giga\hertz}}.

@@

options is given in~\cref{fig:bratu-options}. Results can be reproduced using the FreeFEM script at \url{https://github.com/prj-/jolivet2020petsc}

>>>>

The paper is divided in two main parts. \Cref{sec:KSPHPDDM} introduces
\pk{KSPHPDDM}. The readiness of PETSc for interacting with block Krylov methods
is discussed.  Multiple applications are provided, two of which are within
eigensolver frameworks, showing the interoperability of the proposed
methodology.  \Cref{sec:PCHPDDM} introduces \pk{PCHPDDM}. Again, diverse
problems are dealt with, for example in the context of a nonlinear solver.
Eventually, concluding remarks may be found \cref{sec:conclusion}. All PETSc
keywords and options are typeset in \texttt{typewriter} font, these are
public and documented at \url{https://www.mcs.anl.gov/petsc/petsc-master/docs}.
Codes not already available in other public repositories are available at 
\url{https://github.com/prj-/jolivet2020petsc}.

@@

\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is automatically
registered since PETSc version 3.12 when configuring PETSc with the extra
flag \pk{-{}-download-hpddm}. HPDDM linear solvers can be selected using the command
line option \pk{-ksp\_type hpddm}, or the routine \pk{KSPSetType}(ksp,
\pk{KSPHPDDM}), and the following Krylov methods can be accessed:
\begin{itemize}
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible}, with deflation at each restart;
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown}, with deflation at each iteration;
    \item block GCRODR or flexible GCRODR, with deflation at each restart.
\end{itemize}
While being mathematically equivalent to
their ``standard'' counterparts, the pseudo-block variants try to fuse together multiple similar
operations like matrix--vector
products to achieve higher arithmetic intensity, or to decrease the
number of global synchronizations needed for scalar products.
HPDDM Krylov methods may be selected using the % {\color{red} TODO API FOR KSPHPDDMSetVariant?}
options \pk{-ksp\_hpddm\_type}
\pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)}, or the
routine \pk{KSPHPDDMSetType}(ksp, \pk{KSPHPDDMType}).
Preconditioning variants can be specified via
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether the
preconditioner is applied on the left, on the right, or if it cannot be
represented as a linear operator. In this latter case, the preconditioner is
always applied on the right, except for FCG. HPDDM CG
and its variants BCG and BFBCG only handle left-preconditioning.  It is also
possible to set the preconditioning side through the more common PETSc option
\pk{-ksp\_pc\_side (left$|$right)}.  Finally, convergence monitoring with the
\pk{KSPMonitor} interface is fully supported, as well as the specification of
customized convergence testing via the \pk{KSPSetConvergenceTest} callback.

According to the standard PETSc notation, ksp($A$, $P_{A'}$)
defines a \pk{KSP} used to solve a system $Ax=b$ using a preconditioner $P_{A'}$ built
using an operator $A'$ which is usually, but not necessarily, equal to $A$. While solving
the linear system via \pk{KSPSolve}(ksp, $b$, $x$) or \pk{KSPMatSolve}(ksp, $B$, $X$),
HPDDM will repeatedly call the following PETSc routines:
\begin{itemize}
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, {\small \pk{MAT\_REUSE\_MATRIX}}, {\small
        \pk{PETSC\_DEFAULT}}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}($P_{A'}$, $x$, $y$) for $y = P_{A'}x$;
    \item \pk{PCMatApply}($P_{A'}$, $X$, $Y$) for $Y = P_{A'}X$;
\end{itemize}
As of version 3.14.0, PETSc provides unified routines for applying $A$ or $P_{A'}$
to a tall-and-skinny dense matrix $X$. However, when no specialized implementation
is available, PETSc will loop over the columns of $X$ and successively apply $A$
or $P_{A'}$ to get all columns of $Y$. For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}:
\begin{itemize}
    \item \pk{MatSeqAIJ}, standard sequential sparse matrix, based on compressed sparse row format;
    \item \pk{MatSeqBAIJ}, block sparse matrix, based on block compressed sparse row format;
    \item \pk{MatSeqSBAIJ}, symmetric block sparse matrix stored in upper triangular form.
    \item \pk{MatMPIAIJ}, standard parallel sparse matrix;
    \item \pk{MatShell}, user-defined matrix;
    \item \pk{MatNest}, block-defined matrix with nested submatrices;
    \item \pk{MatSeqAIJCUSPARSE}, standard sequential sparse matrix, offloaded to a GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}
The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    \item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: embedded Krylov method inside $P_{A'}$;
    \item \pk{PCMAT}: multiplication by $A'$, i.e., $P_{A'} = {A'}$;
    \item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    % \item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}~\cite{10.1145/3232850} {\color{red}more info?};
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization,
        e.g., $P_{A'} = {U}^{-1}{L}^{-1}$;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like preconditioners 
\cref{it:hpddm,it:asm,it:bjacobi} is that they most often rely on exact or inexact
factorizations~\cref{it:lu,it:ilu} as subdomain solvers.
In these cases, it is possible to access the so-called local or distributed factored matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to take advantage
of blocked forward eliminations and backward substitutions from the various factorization packages
interfaced with PETSc. In the case of a sparse coefficient matrix $A'$,
this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL
PARDISO or CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense coefficient
matrix $A'$: ScaLAPACK and Elemental~\cite{poulson2013elemental}. 
Future work will consider extending the various implementations of PETSc
preconditioning infrastructure to increase the arithmetic intensity of the
block preconditioner application phase.

The rest of the operations are then performed directly inside HPDDM.
Specifically, within Krylov methods using the Arnoldi process, or when recycling is requested,
one has to compute $QR$ factorizations of tall-and-skinny dense matrices to orthonormalize
candidate basis vectors. Such operations are performed using the CholQR algorithm~\cite{stathopoulos2002block}, or
via the (modified) Gram--Schmidt method.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.
These orthonormalization variants can be selected at runtime with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}. % {\color{red} API?}

@@

With these added functions, it is now possible to do code refactorization for
applications that deal with multiple right-hand sides. This is the case of some
eigensolvers in SLEPc~\cite{hernandez2005ssf}, the Scalable Library for
Eigenvalue Problem computations. SLEPc is an add-on library that extends PETSc
with several classes for linear and nonlinear eigenvalue problems, as well as
tools to, e.g., facilitate the use of shift-and-invert spectral transformations.
The implementations of LOBPCG~\cite{knyazev2001toward} and CISS~\cite{sakurai2003projection}
in SLEPc were previously not using blocking when applying the preconditioner or solving linear systems.
In version 3.14.0, we have adapted these solvers to employ \pk{KSPMatSolve} and \pk{PCMatApply}.
In the case of LOBPCG, which is a purely blocked method, all steps were already
implemented as block operations, except the applications of the preconditioner,
which ruined the benefits of blocking. With the new implementation, the gains
are striking, as reported in~\cref{sec:lobpcg}.

@@

A similar speedup can be observed in terms of runtime, with the time spent for the solution of the eigenvalue problem being reduced from \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1509}{\sec} to \SI[round-mode = places,round-precision=1,group-minimum-digits=3,group-separator = {,}]{1052}{\sec}. These timings, and all which follow except in~\cref{sec:primitives}, have been obtained on {Ir\`ene}, a system composed of \pgfmathprintnumber[assume math mode=true]{1656} nodes with two 24-core Intel Xeon Platinum 8168 clocked at \SI{2.7}{\giga\hertz}.

@@

options is given in~\cref{fig:bratu-options}. Results can be reproduced using the FreeFEM script at \url{https://github.com/prj-/jolivet2020petsc}.
